---
title: "Parse SyRi into R and plot with gggenomes"
author: "Niklas Schandry"
output:
  github_document:
    html_preview: false
---

# About

Here I provide a set of function to read in SyRi-output, and plot it. Overall, the results can look similar to what [`plotsr`](https://github.com/schneebergerlab/plotsr/) creates. The SyRi outputs from [`nf-plotsv`](https://github.com/nschan/nf-plotsv) can be visualized using these functions.
The files included in `data/` for demonstration are the [`plotsr`](https://github.com/schneebergerlab/plotsr/) example files.

Running this requires 'tidyverse' (`dplyr`, `dtplyr`, `magrittr`, and `vroom`) and the output is designed to be compatible with [`gggenomes`](https://github.com/thackl/gggenomes) for plotting. This repo also comes with a snapshot that can be used with `renv::restore()`. The calculation of polygons to draw curves between sequences is directly lifted from the amazing [`GENESPACE`](https://github.com/jtlovell/GENESPACE) package, but  `GENESPACE` is not a dependency.

```{r eval = F}
renv::install("tidyverse","thackl/gggenomes")
```

```{r message = F}
library(tidyverse)
library(gggenomes)
library(magrittr)
source("functions/parse_syri.R") # Contains syri_plot_fills
```

# Input

The functions here are intended to work with the outputs from [`nf-plotsv`](https://github.com/nschan/nf-plotsv). Therefore, the script expects the syri output to be named `genomeA_on_genomeB.syri.out`, and will split based on this. There is _no_ flexibility here.

# Running

In this example, genomeA is col and genomeB is ler.

```{r}
dat <- parse_syri("data/col_on_ler.syri.out",
                  order = data.frame(bin_id = c("col","ler"))
                  )
```

# Plotting

After parsing the data, it is ready for plotting with `gggenomes`.

## gggenomes links

The parsed data can be used with standard `gggenomes` geoms, such as `geom_seq`, `geom_bin`, `geom_link`, etc.

```{r, fig.retina=T, dev='ragg_png', fig.width=12, fig.height=5}
gggenomes::gggenomes(seqs = dat$seqs,
                     links = dat$links) + 
  geom_seq(linewidth = 1) + 
  geom_bin_label(size=7) +
  geom_link(aes(fill = type),color = NA) +
  syri_plot_fills  +
  ggtitle("Synteny between Col and Ler")
```

This is probably ok for the comparison of small-ish regions, but not extremely appealing for whole genomes.

## With polygons

`gggenomes::geom_link()` currently draws simple rectangles. An alternative is to draw links using polygons. These polygons are computed during parsing (unless `no_polygons` is set to `TRUE`). 

```{r, fig.retina=T, dev='ragg_png', fig.width=12, fig.height=5}
gggenomes::gggenomes(seqs = dat$seqs,
                     links = dat$links) + 
  geom_polygon(
    data = dat$polys %>% filter(direct) %>% filter(type == "SYN"),
    aes(
      x = x,
      y = y,
      fill = type,
      group = link_grp
    ),
    alpha = 0.6
  ) +
  geom_polygon(
    data = dat$polys %>% filter(direct) %>% filter(type != "SYN"),
    aes(
      x = x,
      y = y,
      fill = type,
      group = link_grp
    ),
    alpha = 0.8
  ) +
  geom_seq(linewidth = 1) + 
  geom_bin_label(size=7) +
  #geom_link() +
  syri_plot_fills  +
  ggtitle("Synteny between Col and Ler")
```

## Options

### Selecting chromosomes

Sometimes, only a subset of chromosomes is relevant. `parse_syri()` expects chromosome names to be identical across genomes. If that is the case, chromosomes can be selected with the `chroms` parameter

```{r}
dat <- parse_syri("data/col_on_ler.syri.out",
                  order = data.frame(bin_id = c("col","ler")),
                  chroms = c("Chr1","Chr3")
                  )
```

```{r, fig.retina=T, dev='ragg_png', fig.width=12, fig.height=5}
gggenomes::gggenomes(seqs = dat$seqs,
                     links = dat$links) + 
  geom_polygon(
    data = dat$polys %>% filter(direct) %>% filter(type == "SYN"),
    aes(
      x = x,
      y = y,
      fill = type,
      group = link_grp
    ),
    alpha = 0.6
  ) +
  geom_polygon(
    data = dat$polys %>% filter(direct) %>% filter(type != "SYN"),
    aes(
      x = x,
      y = y,
      fill = type,
      group = link_grp
    ),
    alpha = 0.8
  ) +
  geom_seq(linewidth = 1) + 
  geom_bin_label(size=7) +
  #geom_link() +
  syri_plot_fills  +
  ggtitle("Synteny between Col and Ler Chromosomes 1 and 3")
```

### Spacing

Sometimes, the default spacing between chromosomes may not be optimal. `parse_syri()` follows gggenomes in spacing rules. If spacing is < 1, it is relative to the longest bin / sqrt(number of sequences), if it is >= 1 it is base pairs. The default is 0.05 (as for gggenomes)

#### In basepairs

```{r}
dat <- parse_syri("data/col_on_ler.syri.out",
                  order = data.frame(bin_id = c("col","ler")),
                  spacing = 5000000 # spacing in bp
                  )
```

Of course, if the spacing was changed, this also needs to be adjusted in gggenomes:

```{r, fig.retina=T, dev='ragg_png', fig.width=12, fig.height=5}
gggenomes::gggenomes(seqs = dat$seqs,
                     links = dat$links,
                     spacing = 5000000) + 
  geom_polygon(
    data = dat$polys %>% filter(direct) %>% filter(type == "SYN"),
    aes(
      x = x,
      y = y,
      fill = type,
      group = link_grp
    ),
    alpha = 0.6
  ) +
  geom_polygon(
    data = dat$polys %>% filter(direct) %>% filter(type != "SYN"),
    aes(
      x = x,
      y = y,
      fill = type,
      group = link_grp
    ),
    alpha = 0.8
  ) +
  geom_seq(linewidth = 1) + 
  geom_bin_label(size=7) +
  syri_plot_fills  +
  ggtitle("Synteny between Col - Ler with 5MB spacing between chromsomes")
```

#### Relative

4 times the standard spacing:

```{r}
dat <- parse_syri("data/col_on_ler.syri.out",
                  order = data.frame(bin_id = c("col","ler")),
                  spacing = 0.2 # relative spacing
                  )
```

```{r, fig.retina=T, dev='ragg_png', fig.width=12, fig.height=5}
gggenomes::gggenomes(seqs = dat$seqs,
                     links = dat$links,
                     spacing = 0.2) + 
  geom_polygon(
    data = dat$polys %>% filter(direct) %>% filter(type == "SYN"),
    aes(
      x = x,
      y = y,
      fill = type,
      group = link_grp
    ),
    alpha = 0.6
  ) +
  geom_polygon(
    data = dat$polys %>% filter(direct) %>% filter(type != "SYN"),
    aes(
      x = x,
      y = y,
      fill = type,
      group = link_grp
    ),
    alpha = 0.8
  ) +
  geom_seq(linewidth = 1) + 
  geom_bin_label(size=7) +
  syri_plot_fills  +
  ggtitle("Synteny between Col - Ler, spacing increased 4x")
```

### No resizing

By default, short syntenic regions larger than 5000 bp are resized to make them visible. Since this does not reflect the original input, this can be disabled:

```{r}
dat <- parse_syri("data/col_on_ler.syri.out",
                  order = data.frame(bin_id = c("col","ler")),
                  resize_polygons = F)
```

```{r, fig.retina=T, dev='ragg_png', fig.width=12, fig.height=5}
gggenomes::gggenomes(seqs = dat$seqs,
                     links = dat$links) + 
  geom_polygon(
    data = dat$polys %>% filter(direct) %>% filter(type == "SYN"),
    aes(
      x = x,
      y = y,
      fill = type,
      group = link_grp
    ),
    alpha = 0.6
  ) +
  geom_polygon(
    data = dat$polys %>% filter(direct) %>% filter(type != "SYN"),
    aes(
      x = x,
      y = y,
      fill = type,
      group = link_grp
    ),
    alpha = 0.8
  ) +
  geom_seq(linewidth = 1) + 
  geom_bin_label(size=7) +
  syri_plot_fills  +
  ggtitle("Synteny between Col and Ler without resizing")
```

### Minimum resize size 

Only regions larger than `min_polygon_feat_size` are resized (default 5000), this can be modified to also include smaller regions

```{r}
dat <- parse_syri("data/col_on_ler.syri.out",
                  order = data.frame(bin_id = c("col","ler")),
                  resize_polygons = T,
                  min_polygon_feat_size = 1000)
```

Naturally, this will create a busier plot.

```{r, fig.retina=T, dev='ragg_png', fig.width=12, fig.height=5}
gggenomes::gggenomes(seqs = dat$seqs,
                     links = dat$links) + 
  geom_polygon(
    data = dat$polys %>% filter(direct) %>% filter(type == "SYN"),
    aes(
      x = x,
      y = y,
      fill = type,
      group = link_grp
    ),
    alpha = 0.6
  ) +
  geom_polygon(
    data = dat$polys %>% filter(direct) %>% filter(type != "SYN"),
    aes(
      x = x,
      y = y,
      fill = type,
      group = link_grp
    ),
    alpha = 0.8
  ) +
  geom_seq(linewidth = 1) + 
  geom_bin_label(size=7) +
  syri_plot_fills  +
  ggtitle("Synteny between Col and Ler, resizing regions larger than 999bp")
```

### Resize output size

Regions are resized to have a certain length relative to the chromosome, controlled by `resize_polygons_size`, which defaults to `0.003` (0.3%) of the chromosome length.
Altering this will make resized regions larger, or smaller.

```{r}
dat <- parse_syri("data/col_on_ler.syri.out",
                  order = data.frame(bin_id = c("col","ler")),
                  resize_polygons = T,
                  resize_polygons_size = 0.01)
```

This will produce wider polygons for resized links.

```{r, fig.retina=T, dev='ragg_png', fig.width=12, fig.height=5}
gggenomes::gggenomes(seqs = dat$seqs,
                     links = dat$links) + 
  geom_polygon(
    data = dat$polys %>% filter(direct) %>% filter(type == "SYN"),
    aes(
      x = x,
      y = y,
      fill = type,
      group = link_grp
    ),
    alpha = 0.6
  ) +
  geom_polygon(
    data = dat$polys %>% filter(direct) %>% filter(type != "SYN"),
    aes(
      x = x,
      y = y,
      fill = type,
      group = link_grp
    ),
    alpha = 0.8
  ) +
  geom_seq(linewidth = 1) + 
  geom_bin_label(size=7) +
  syri_plot_fills  +
  ggtitle("Synteny between Col and Ler")
```

# Multiple genomes

Comparing two genomes is nice, but more are better.

`parse_syri` can handle multiple outputs in one go:

```{r}
file_list <- list.files("data", full.names = T)
syri_order <- data.frame(bin_id = c("col", "ler", "cvi", "eri"))
dat <- parse_syri(file_list, order = syri_order)
```

Making a plot from this works the same way of making a plot of only one comparison. The order of sequences is set via the `order` argument to `parse_syri()`

```{r, fig.retina=T, dev='ragg_png', fig.width=12, fig.height=5}
gggenomes::gggenomes(seqs = dat$seqs,
                     links = dat$links) + 
  geom_polygon(
    data = dat$polys %>% filter(direct) %>% filter(type == "SYN"),
    aes(
      x = x,
      y = y,
      fill = type,
      group = link_grp
    ),
    alpha = 0.6
  ) +
  geom_polygon(
    data = dat$polys %>% filter(direct) %>% filter(type != "SYN"),
    aes(
      x = x,
      y = y,
      fill = type,
      group = link_grp
    ),
    alpha = 0.8
  ) +
  geom_seq(linewidth = 1) + 
  geom_bin_label(size=7) +
  syri_plot_fills  +
  ggtitle("Synteny between Col - Ler - Cvi - Eri")
```

# Keeping chromosomes separate

By default, `parse_syri()` takes all chromosomes from the same genome (`bin_id`) and puts them on one axis, adding space between them as needed (See [spacing](#Spacing)).

Sometimes, it might be useful to have the chromosomes each on their own coordinate system instead. This can be done by making use of the `chroms` argument to read each chromosome individually and then combining them. Below is an example for the col-ler-cvi-eri data used above and included in `data/`.

```{r}
file_list <- list.files("data", full.names = T)
syri_order <- data.frame(bin_id = c("col", "ler", "cvi", "eri"))
chromosomes <-  c("Chr1", "Chr2", "Chr3", "Chr4", "Chr5")
dat_tmp <- lapply(chromosomes,
                  \(chrom) parse_syri(file_list, order = syri_order, chroms = chrom))

# Bind sequences
dat$seqs <- lapply(1:length(chromosomes), \(l) pluck(dat_tmp, l, "seqs")) %>%
  bind_rows()
# Create y coordinates for sequences
seq_pos <- left_join(dat$seqs, syri_order %>%
                       mutate(y = rev(1:length(bin_id))),
                     by = join_by(bin_id))
# Bind links
dat$links <- lapply(1:length(chromosomes), \(l) pluck(dat_tmp, l, "links")) %>%
  bind_rows()
# Bind polygons
dat$polys <- lapply(1:length(chromosomes), \(l) pluck(dat_tmp, l, "polys")) %>%
  bind_rows()
# Add seq_id column to polygons, only keep polygons that connect the same chromosome
dat$polys <- dat$polys %>%
  mutate(
    Chr_grp1 = str_extract_all(link, "Chr[0-9]*", simplify = T)[, 1],
    Chr_grp2 = str_extract_all(link, "Chr[0-9]*", simplify = T)[, 2]
  ) %>%
  filter(Chr_grp1 == Chr_grp2) %>%
  mutate(seq_id = Chr_grp1)
```

This can then be plotted. Note that `geom_segment()` should be used to draw chromosomes, since `geom_seq()` would again place the chromosomes onto a single axis.

```{r chrom_plot, fig.retina=T, dev='ragg_png', fig.width = 12, fig.height = 24}
gggenomes::gggenomes(seqs = dat$seqs,
                     links = dat$links) + 
  geom_polygon(
    data = dat$polys %>% filter(direct) %>% filter(type == "SYN"),
    aes(
      x = x,
      y = y,
      fill = type,
      group = link_grp
    ),
    alpha = 0.6
  ) +
  geom_polygon(
    data = dat$polys %>% filter(direct) %>% filter(type != "SYN"),
    aes(
      x = x,
      y = y,
      fill = type,
      group = link_grp
    ),
    alpha = 0.8
  ) +
  geom_segment(aes(x = 0, xend = length, y=y, yend=y), data = seq_pos) +
  geom_bin_label(size=7, 
                 # Avoid overly long extension of x to the left
                 expand_left = 1e-2,
                 nudge_left = 5e-3) +
  facet_wrap(~seq_id, ncol = 1,  scales = "free_x") +
  syri_plot_fills +
  theme(strip.background = element_rect(fill = "white")) +
  ggtitle("Synteny between Col - Ler - Cvi - Eri")
```


# Contributing

If you encounter any problems, please open an issue. If you have suggestions for improvement, please open a pull request.